Co-moving stars in \textsl{Gaia DR1}

### 2016-10-06 paper/ms.tex
author stuff
reworking some of the methods

% define macros for text
\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\gaia}{\project{Gaia}}
\newcommand{\rave}{\project{\acronym{RAVE}}}
\newcommand{\apogee}{\project{\acronym{APOGEE}}}
\newcommand{\documentname}{\textsl{Article}}
\newcommand{\sectionname}{Section}
\newcommand{\figname}{Figure}
\newcommand{\eqname}{Equation}
\newcommand{\dr}{\acronym{DR1}}

% define macros for math
\newcommand{\given}{\,|\,}
\newcommand{\normal}{{\mathcal{N}}}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\transp}[1]{{#1}^{\!\mathsf{T}}}
\newcommand{\inv}[1]{{#1}^{-1}}
\newcommand{\bs}[1]{\boldsymbol{#1}}
\newcommand{\vperp}{\bs{v}^\perp}
\newcommand{\propm}{\bs{\mu}}
\newcommand{\matrx}[1]{\mathbf{#1}}
\newcommand{\kms}{\rm km~s^{-1}}
\newcommand{\data}{\mathrm{data}}
\newcommand{\snr}{[S/N]_\varpi}

@@

\title{Co-moving stars in \textsl{Gaia DR1}}
\author{Adrian M. Price-Whelan\altaffilmark{\pu,\adrn},
        Semyeong Oh\altaffilmark{\pu},
        David W. Hogg\altaffilmark{\ccpp,\mpia},
        Timothy D. Morton\altaffilmark{\pu},
        David N. Spergel\altaffilmark{\pu,\cca}
}

% Affiliations
\newcommand{\pu}{1}
\newcommand{\adrn}{2}
\newcommand{\ccpp}{3}
\newcommand{\mpia}{4}
\newcommand{\cca}{5}

\altaffiltext{\pu}{Department of Astrophysical Sciences,
                   Princeton University, Princeton, NJ 08544, USA}
\altaffiltext{\adrn}{To whom correspondence should be addressed:
                     \texttt{adrn@princeton.edu}}
\altaffiltext{\ccpp}{Center for Cosmology and Particle Physics,
                     Department of Physics,
                     New York University, 4 Washington Place,
                     New York, NY 10003, USA}
\altaffiltext{\mpia}{Max-Planck-Institut f\"ur Astronomie,
                     K\"onigstuhl 17, D-69117 Heidelberg, Germany}
\altaffiltext{\cca}{Simons Center for Computational Astrophysics, ...,
                    New York, NY XXXXX, USA}

@@

The primary data set used in this \documentname\ is the Tycho-Gaia Astrometric
Solution (TGAS), released as a part of Data Release 1 of the Gaia mission
(\citealt{many}). TGAS contains astrometric measurements (sky position,
parallax, and proper motions) and associated covariance matrices for a large
fraction of the \project{Tycho-2} catalog (\citealt{tycho2}) with median
astrometric precision equivalent to the \project{Hipparcos} catalog ($\approx
0.3~{\rm mas}$; \citealt{hipparcos}). In terms of parallax signal-to-noise
($\snr = \varpi/\sigma_\varpi$), the TGAS catalog contains 1240114 stars with
$\snr > 4$, 619618 stars with $\snr > 8$, and 197105 stars with $\snr > 16$. The
increase in precision over Tycho-2 and increase in volume over Hipparcos makes
the TGAS catalog a powerful data set to search for widely-separated, co-moving
stars.

In the sections that follow, we construct a statistical model that propagates
the non-trivial uncertainties in the data to our beliefs about the likelihood
that a given pair are co-moving. Here, we instead compute quantities directly
from the astrometric measurements to visualize and emphasize the quality of the
data. For each star with $\snr > 16$, we use the reported parallax and parallax
error to compute a distance corrected for the Lutz-Kelker bias
(\citealt{lutzkelker}):
\begin{equation}
  d_{\rm LK} = 1000 \, \left[\frac{\varpi}{2} \,
    \left(1 + \sqrt{1 - \frac{16}{\snr^2}} \right) \right]^{-1} \, {\rm pc}
\end{equation}
We use this distance and the sky positions of each source to compute 3D
Cartesian positions, $\boldsymbol{x}$, for all stars and find the nearest
neighbor to each source in 3-space using the KDTree implementation in
\project{scikit-learn} (\citealt{scikit-learn}). We remove duplicates from
this sample and are left with 138962 pairs of stars. For all pairs, we then
compute the tangential velocity difference
\begin{equation}
  |\Delta \bs{v}_{\rm t}| = d_{\rm LK} \,
                            \sqrt{(\mu_{\alpha,1}-\mu_{\alpha,2})^2 +
                                  (\mu_{\delta,1}-\mu_{\delta,2})^2}
\end{equation}
\figname~\ref{fig:dv-sep} shows these quantities plotted for all pairs that pass
the above signal-to-noise cut. Pairs with small physical separation also tend to
have small velocity differences. Intriguingly, there appear to be an abundance
of pairs with small velocity separation at separations $|\delta\bs{x}| \gtrsim
1~{\rm pc}$.

@@

For a given pair of stars $(i,j)$, we compute the fully marginalized likelihood
(FML) that the pair has the same 3-space velocity (hypothesis 1,
$\mathcal{L}_1$) and the FML of the stars having different 3-space velocities
(hypothesis 2, $\mathcal{L}_2$). We use the ratio of these FMLs as a scalar for
selecting candidate widely-separated binary stars. \gaia\ \dr\ only contains
parallax, $\varpi$, and two proper motion components, $\propm = (\mu_\alpha,
\mu_\delta)$ for each star, so for the majority of sources we only have
information about the tangential velocities of the stars. Where there is overlap
with \rave\ or \apogee, some number of observed radial velocities, $v_r$, with
observed variances $\sigma^2_{v_r}$ (independent of the \gaia\ data) may also be
available. We assume that the uncertainties in these observed quantites are
known with covariance matrix $\matrx{C}$. The likelihood functions $L_1, L_2$
are marginalized over true 3-space velocity and distance for each star in the
pair.
\begin{align}
  \mathcal{L}_1 &=
    \int \, \dd d_i \, \dd d_j \, \dd^3 \bs{v} \,
    L_i(\bs{v}, d_i) \,
    L_j(\bs{v}, d_j) \,
    p(\bs{v}) \, p(d_i \given \varpi_i) \, p(d_j \given \varpi_j) \\
  \mathcal{L}_2 &=
    \int \, \dd d_i \, \dd d_j \, \dd^3 \bs{v}_i \, \dd^3 \bs{v}_j \,
    L_i(\bs{v}_i, d_i) \,
    L_j(\bs{v}_j, d_j) \,
    p(\bs{v}_i) \, p(\bs{v}_j) \, p(d_i \given \varpi_i) \, p(d_j \given \varpi_j) \label{eq:hyp2}
\end{align}
where
\begin{align}
  L(\bs{v}, d) &=
    \left[\det\left(\frac{\matrx{C}^{-1}}{2\pi}\right)\right]^{1/2} \,
    \exp \left[ -\frac{1}{2} \transp{\left(\bs{x} - \bs{x}_\theta \right)} \,
    \matrx{C}^{-1} \,
    \left(\bs{x} - \bs{x}_\theta \right) \right] \\
  \bs{x} &= \transp{(\begin{array}{c c c} \mu_\alpha &
    \mu_\delta & v_r \end{array})} \\
  \bs{x}_\theta &= \transp{\left(\begin{array}{c c c} d^{-1}\,v_\alpha &
    d^{-1}\,v_\delta & \tilde{v}_r \end{array}\right)}
\end{align}
and we compute the posterior pdfs for the true distances, $p(d\given\varpi)$,
using a uniform space density prior \citep{Astraatmadja:2016} on heliocentric
distance
\begin{equation}
  p(d) \propto d^2
\end{equation}
We assume Gaussian priors on the true velocity components
\begin{equation}
  p(\bs{v}) = \normal(0,25)~\kms
\end{equation}

>>>>

% define macros for text
\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\gaia}{\project{Gaia}}
\newcommand{\rave}{\project{\acronym{RAVE}}}
\newcommand{\apogee}{\project{\acronym{APOGEE}}}
\newcommand{\documentname}{\textsl{Article}}
\newcommand{\sectionname}{Section}
\newcommand{\figname}{Figure}
\newcommand{\eqname}{Equation}
\newcommand{\dr}{\acronym{DR1}}
\newcommand{\tgas}{\acronym{TGAS}}

% define macros for math
\newcommand{\given}{\,|\,}
\newcommand{\normal}{{\mathcal{N}}}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\transp}[1]{{#1}^{\!\mathsf{T}}}
\newcommand{\inv}[1]{{#1}^{-1}}
\newcommand{\bs}[1]{\boldsymbol{#1}}
\newcommand{\vperp}{\bs{v}^\perp}
\newcommand{\propm}{\bs{\mu}}
\newcommand{\matrx}[1]{\mathbf{#1}}
\newcommand{\kms}{\rm km~s^{-1}}
\newcommand{\msun}{{\rm M}_\odot}
\newcommand{\data}{\mathrm{data}}
\newcommand{\snr}{[S/N]_\varpi}

@@

\title{Co-moving stars in \textsl{Gaia DR1}}
\author{Semyeong Oh\altaffilmark{\pu,\lead},
        Adrian M. Price-Whelan\altaffilmark{\pu},
        David W. Hogg\altaffilmark{\ccpp,\mpia},
        Timothy D. Morton\altaffilmark{\pu},
        David N. Spergel\altaffilmark{\pu,\cca}
}

% Affiliations
\newcommand{\pu}{1}
\newcommand{\lead}{2}
\newcommand{\ccpp}{3}
\newcommand{\mpia}{4}
\newcommand{\cca}{5}

\altaffiltext{\pu}{Department of Astrophysical Sciences,
                   Princeton University, Princeton, NJ 08544, USA}
\altaffiltext{\lead}{To whom correspondence should be addressed:
                     \texttt{semyeong@astro.princeton.edu}}
\altaffiltext{\ccpp}{Center for Cosmology and Particle Physics,
                     Department of Physics,
                     New York University, 4 Washington Place,
                     New York, NY 10003, USA}
\altaffiltext{\mpia}{Max-Planck-Institut f\"ur Astronomie,
                     K\"onigstuhl 17, D-69117 Heidelberg, Germany}
\altaffiltext{\cca}{Simons Center for Computational Astrophysics, ...,
                    New York, NY XXXXX, USA}

@@

The primary data set used in this \documentname\ is the Tycho-Gaia Astrometric
Solution (\tgas), released as a part of Data Release 1 (\dr) of the Gaia mission
(\citealt{many}). \tgas\ contains astrometric measurements (sky position,
parallax, and proper motions) and associated covariance matrices for a large
fraction of the \project{Tycho-2} catalog (\citealt{tycho2}) with median
astrometric precision comparable to that of the \project{Hipparcos} catalog
($\approx 0.3~{\rm mas}$; \citealt{hipparcos}). In terms of parallax
signal-to-noise ($\snr = \varpi/\sigma_\varpi$), the \tgas\ catalog contains
42385 high-precision stars with $\snr > 32$: here we visualize quantities
computed from this high-precsision sample to emphasize the quality of the data.
In the sections that follow, we construct a statistical model that propagates
the non-trivial uncertainties in the data to our beliefs about the likelihood
that a given pair are co-moving.

For each star with $\snr > 32$, we use the reported parallax and parallax
error to compute a distance corrected for the Lutz-Kelker bias
(\citealt{lutzkelker}):
\begin{equation}
  d_{\rm LK} = 1000 \, \left[\frac{\varpi}{2} \,
    \left(1 + \sqrt{1 - \frac{16}{\snr^2}} \right) \right]^{-1} \, {\rm pc}
\end{equation}
We use this distance and the sky positions of each source to compute 3D
Cartesian positions, $\boldsymbol{x}$, for all stars and find the nearest
neighbor to each source in 3-space using the KDTree implementation in
\project{scikit-learn} (\citealt{scikit-learn}). We remove duplicates from
this sample and are left with 29870 pairs of stars. For all pairs, we then
compute the tangential velocity difference
\begin{equation}
  |\Delta \bs{v}_{\rm t}| = d_{\rm LK} \,
                            \sqrt{(\mu_{\alpha,1}-\mu_{\alpha,2})^2 +
                                  (\mu_{\delta,1}-\mu_{\delta,2})^2}
\end{equation}
\figname~\ref{fig:dv-sep} shows the tangential velocity difference against the
physical separation plotted for all pairs that pass the above signal-to-noise
cut. Pairs with small physical separation also tend to have small velocity
differences. Intriguingly, there appear to be an abundance of pairs with small
velocity separation at separations $|\delta\bs{x}| \gtrsim 1~{\rm pc}$.

@@

The abundance of pairs of stars with small velocity difference in the
high-precsision sample of \figname~\ref{fig:dv-sep} suggests that there are a
significant number of co-moving stars in the \tgas\ data at separations as large
as a few parsecs. Here we develop a method to select high-confidence co-moving
stars that properly incorporates the complex uncertainties associated with the
\gaia\ data. We make the following assumptions in order to construct a
statistical model (a likelihood function with explicit priors on our
parameters):
\begin{itemize}
  \item We assume that the uncertainties in the data---parallax, $\varpi$, and
    two proper motion components, $\propm = (\begin{array}[t]{c c} \mu_\alpha &
    \mu_\delta\end{array})^\mathsf{T}$---are Gaussian with known covariances.
  \item We assume that the tangential velocities of a given pair of stars in the
    \tgas\ sample (relative to the solar system barycenter) are either (1)
    co-moving with velocities equal to the circular velocity of a $2~\msun$
    binary star system with the given physical separation, (2) co-moving with
    velocities drawn from a zero-mean Gaussian with velocity dispersion $\sigma
    = 1~\kms$, or (3) drawn from a mixture of three isotropic, zero-mean
    Gaussian distributions with velocity dispersions equal to $(\sigma_1,
    \sigma_2, \sigma_3) = (8, 32, 128)~\kms$ meant to represent young disk
    stars, old disk stars, and halo stars.
\end{itemize}

Under these assumptions, the likelihood function for the proper-motion data for
a star given true distance, $d$, and true tangential velocity, $\bs{v}^\perp =
(\begin{array}[t]{c c} v_\alpha & v_\delta\end{array})^\mathsf{T}$ is
\begin{align}
  L(\bs{v}^\perp, d) &=
    \left[\det\left(\frac{\matrx{C}^{-1}}{2\pi}\right)\right]^{1/2} \,
    \exp \left[ -\frac{1}{2} \transp{\left(\bs{\mu} - \bs{x}_\theta \right)} \,
    \matrx{C}^{-1} \,
    \left(\bs{\mu} - \bs{x}_\theta \right) \right] \\
  \bs{x}_\theta &= d^{-1} \, \bs{v}^\perp
\end{align}

\todo{APW is working on this text...}
We compute the fully
marginalized likelihood (FML) that the pair has the same 3-space velocity
(hypothesis 1, $\mathcal{L}_1$) and the FML of the stars having different
3-space velocities (hypothesis 2, $\mathcal{L}_2$). We use the ratio of these
FMLs as a scalar for selecting candidate co-moving stars, as described below in
more detail. The likelihood functions $L_1, L_2$ are marginalized over true
3-space velocity and distance for each star in the pair.
\begin{align}
  \mathcal{L}_1 &=
    \int \, \dd d_i \, \dd d_j \, \dd^3 \bs{v} \,
    L_i(\bs{v}, d_i) \,
    L_j(\bs{v}, d_j) \,
    p(\bs{v}) \, p(d_i \given \varpi_i) \, p(d_j \given \varpi_j) \\
  \mathcal{L}_2 &=
    \int \, \dd d_i \, \dd d_j \, \dd^3 \bs{v}_i \, \dd^3 \bs{v}_j \,
    L_i(\bs{v}_i, d_i) \,
    L_j(\bs{v}_j, d_j) \,
    p(\bs{v}_i) \, p(\bs{v}_j) \, p(d_i \given \varpi_i) \, p(d_j \given \varpi_j) \label{eq:hyp2}
\end{align}
where

and we compute the posterior pdfs for the true distances, $p(d\given\varpi)$,
using a uniform space density prior \citep{Astraatmadja:2016} on heliocentric
distance
\begin{equation}
  p(d) \propto d^2
\end{equation}
We assume Gaussian priors on the true velocity components
\begin{equation}
  p(\bs{v}) = \normal(0,25)~\kms
\end{equation}
